Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-09-08 psy_peds.Rmd
adding time series analysis

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "/4ceData/Input"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
```

@@

@@

## Adding columns with hospitalization number and hospitalization length to the clinical raw file
```{r}
count_sequences_hospitalisation <- function(df, ...) {
  # TODO: To be optimized to make it faster 
  total_span <- data.frame(total_span = seq(min(df$days_since_admission),
                    max(df$days_since_admission))
  )
  total_span <- left_join(total_span, df[,c("in_hospital", "days_since_admission")],
            by = c("total_span" = "days_since_admission")
  ) %>% replace_na(list(in_hospital = 0))
  count_sequences <- rle(total_span$in_hospital)
  count_sequences_1 <- lapply(count_sequences, function(x) x[count_sequences$values == 1])
  n_sequences <- seq_along(count_sequences_1$lengths)
  sequences <- rep.int(n_sequences, count_sequences_1$lengths)
  sequences_len <- rep.int(count_sequences_1$lengths, count_sequences_1$lengths)
  df[df$in_hospital == 1, "n_hospitalisation"] <- sequences
  df[df$in_hospital == 1, "len_hospitalisation"] <- sequences_len
  return(df)
}
clinical_raw <- clinical_raw %>%
  group_by(patient_num, cohort) %>% 
  group_modify(count_sequences_hospitalisation)
```

@@

patient_count_monthly <- merge( patient_count_per_month_category, patient_count_per_month)
```

```{r}
#estimate the number of patients per week
patient_count_monthly %>%
  dplyr::filter(icd_description_category == "Psy") %>% 
  dplyr::mutate(month = as.Date(month)) %>% 
  dplyr::mutate(period = ifelse( month < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic"), 
                percentage = patients_per_category * 100 / total_patients) %>% 
  ggplot(aes(x = month, y = percentage, fill = period)) +
  #geom_line()+
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  labs(y = "Percentage",
       x = "Calendar Months",
       title = "Percentage of patients with mental health related ICD codes (monthly)") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

@@

>>>>

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "Input"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
```

@@

```{r}
# Adding filtering on in_hospital == 1 here to reduce subsequent compute time
clinical_raw <- filter(clinical_raw, in_hospital == 1)
```

@@

## Adding columns with hospitalization number and hospitalization length to the clinical raw file
```{r}
count_sequences_hospitalisation <- function(df, ...) {
  # TODO: To be optimized
  total_span <- data.frame(total_span = seq(min(df$days_since_admission),
                                            max(df$days_since_admission))
  ) %>% 
    left_join(df[,c("in_hospital", "days_since_admission")],
              by = c("total_span" = "days_since_admission")
    ) %>% replace_na(list(in_hospital = 0))
  count_sequences <- rle(total_span$in_hospital)
  count_sequences_1 <- lapply(count_sequences, function(x) x[count_sequences$values == 1])
  n_sequences <- seq_along(count_sequences_1$lengths)
  sequences <- rep.int(n_sequences, count_sequences_1$lengths)
  sequences_len <- rep.int(count_sequences_1$lengths, count_sequences_1$lengths)
  df[df$in_hospital == 1, "n_hospitalisation"] <- sequences
  df[df$in_hospital == 1, "len_hospitalisation"] <- sequences_len
  return(df)
}
clinical_raw <- clinical_raw %>%
  group_by(patient_num, cohort) %>% 
  group_modify(count_sequences_hospitalisation)
```

@@

patient_count_monthly <- merge( patient_count_per_month_category, patient_count_per_month) %>%
  dplyr::mutate(month = as.Date(month)) %>% 
  dplyr::mutate(period = ifelse( month < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic"), 
                percentage = patients_per_category * 100 / total_patients) 
```

```{r}
#estimate the number of patients per week
patient_count_monthly %>% 
  filter(icd_description_category == "Psy") %>% 
  ggplot(aes(x = month, y = percentage, fill = period)) +
  #geom_line()+
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  labs(y = "Percentage",
       x = "Calendar Months",
       title = "Percentage of patients with mental health related ICD codes (monthly)") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

@@

## Interrupted Time-Serie Analysis

```{r} 
# 
patient_count_monthly_psy <- patient_count_monthly %>% 
  filter(icd_description_category == "Psy", 
         !month %in% c(as.Date("2019-01-01"), as.Date("2019-02-01")) & month != max(month)) %>% 
  mutate(time = seq_along(month), 
         month_year = format(month, "%m"))
```

```{r}
# Model without seasonality
time_serie_analysis_plotting <- function(model, patient_count_monthly_psy) {
  summary(model)
  output_model <- broom::tidy(model)
  coeff_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "estimate"][[1]], 3)
  pvalue_slope <- round(output_model[output_model$term == "periodduring_pandemic:time", "p.value"][[1]], 3)
  annotation <- paste0("Lockdown slope: ", coeff_slope, " (pvalue: ", pvalue_slope, ")")
  pred_model <- predict(model, patient_count_monthly_psy, type = "response", interval = "confidence") %>% as.data.frame()
  pred_model_b <- predict(model, mutate(patient_count_monthly_psy, period = "before_pandemic"), type = "response", interval = "confidence") %>%
    as.data.frame()
  names(pred_model_b) <- paste0(names(pred_model_b), "_wo_intervention")

  pred_model <- bind_cols(patient_count_monthly_psy, pred_model, pred_model_b)
  plot <- ggplot(pred_model, aes(x = month, y = percentage)) + 
    geom_point() +
    geom_line(aes(x = month, y = fit, colour = "Mean pandemic")) +
    geom_line(aes(x = month, y = fit_wo_intervention, colour = "Expected mean w/o pandemic"), linetype = "dashed") + 
    scale_color_manual(name = "Time Series projection", values = c("Mean pandemic" = "black", "Expected mean w/o pandemic" = "red")) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
    # geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
    geom_vline(xintercept = as.Date(pandemic_start_date), linetype = "dashed") +
    annotate("text", x = as.Date("2019-10-01"), y = max(pred_model$percentage) + 5, label = annotation) +
    theme(legend.position = "bottom")
  return(plot)
}
```

```{r}
# Time series analysis of percentage of psy codes per encounter, linear (without seasonality)
ts_linear_percent <- lm(percentage ~ period * time, patient_count_monthly_psy)
time_serie_analysis_plotting(ts_linear_percent, patient_count_monthly_psy) + 
    labs(title = "Interrupted Time-Serie Analysis: Percentage of encounters with psychiatric ICD codes", 
         subtitle = "Without seasonality effect",
         x = "Percentage", 
         y = "Month")
ggsave("output/time_series_percent_linear.png", width = 18, height = 9, units="cm")

```{r}
# Time series analysis of percentage of psy codes per encounter, with seasonality
ts_seasonal_percent <- lm(percentage ~ period*time + harmonic(month_year, 2, 12), patient_count_monthly_psy)
time_serie_analysis_plotting(ts_seasonal_percent, patient_count_monthly_psy) +
  labs(title = "Interrupted Time-Serie Analysis: Percentage of encounters with psychiatric ICD codes", 
       subtitle = "With seasonality effect",
       x = "Percentage", 
       y = "Month")
ggsave("output/time_series_percent_seasonal.png", width = 18, height = 9, units="cm")

```{r}
# Time series analysis of percentage of icd code count, with seasonality
ts_count_seasonal <- glm(patients_per_category ~ period*time + harmonic(month_year, 1, 12),
                       family = quasipoisson,
                       patient_count_monthly_psy)

output_ts_seasonal_count <- broom::tidy(ts_count_seasonal)
coeff_slope <- round(output_ts_seasonal_count[output_ts_seasonal_count$term == "periodduring_pandemic:time", "estimate"][[1]], 3)
pvalue_slope <- round(output_ts_seasonal_count[output_ts_seasonal_count$term == "periodduring_pandemic:time", "p.value"][[1]], 3)
annotation <- paste0("Lockdown slope: ", coeff_slope, " (pvalue: ", pvalue_slope, ")")
pred_ts_seasonal_count <- predict(ts_count_seasonal, patient_count_monthly_psy, type = "response") %>% data.frame(fit = .)
pred_ts_seasonal_count_b <- predict(ts_count_seasonal, mutate(patient_count_monthly_psy, period = "before_pandemic"), type = "response")%>% data.frame(fit = .)
names(pred_ts_seasonal_count_b) <- paste0(names(pred_ts_seasonal_count_b), "_wo_intervention")

pred_ts_seasonal_count <- bind_cols(patient_count_monthly_psy, pred_ts_seasonal_count, pred_ts_seasonal_count_b)
ggplot(pred_ts_seasonal_count, aes(x = month, y = patients_per_category)) + 
  geom_point() +
  geom_line(aes(x = month, y = fit, colour = "Mean pandemic")) +
  geom_line(aes(x = month, y = fit_wo_intervention, colour = "Expected mean w/o pandemic"), linetype = "dashed") + 
  scale_color_manual(name = "Time Series projection", values = c("Mean pandemic" = "black", "Expected mean w/o pandemic" = "red")) +
  geom_vline(xintercept = as.Date(pandemic_start_date), linetype = "dashed") +
  annotate("text", x = as.Date("2019-10-01"), y = max(pred_ts_seasonal_count$patients_per_category) + 5, label = annotation) +
  labs(title = "Interrupted Time-Serie Analysis: Count of encounters with psychiatric ICD codes", 
       subtitle = "With seasonality effect",
       x = "Percentage", 
       y = "Month") + 
  theme(legend.position = "bottom")
ggsave("output/time_series_count_seasonal.png", width = 18, height = 9, units="cm")
```
